\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Type-Preserving Mutations

### 2021-04-17 papers/oopsla-21/main.tex
Moved paragraph to section 2. Fixed figures.

%
%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
%% In this work, we take a close look at \fuzzchick for opportunities to improve.
%
%% Notably, we establish several aspects of \fuzzchick that can be
%% revised.%
%
The approach taken by \fuzzchick leaves considerable room for
improvements in several aspects.%
\footnote{Some of these were already identified by the authors of \fuzzchick in
  their original work.}
%But \fuzzchick suffers several disadvantages.%
%
In particular:
%
\begin{inparaenum}
\item type-preserving mutations, if not done carefully, can become superficial
  and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause large delays if interesting values are
  enqueued frequently and there is no way to prioritize them, and
\item using an heuristic to assign a certain ``mutation energy'' to each
  interesting test case (often referred to as a \emph{power schedule}) requires
  fine tuning and can be hard to generalize to work well under different testing
  scenarios.
\end{inparaenum}
%
Moreover, we replicated the Information-Flow Control (IFC) stack machine case
study bundled with \fuzzchick and observed a surprising limitation:
%
after repeating each experiment 30 times, \fuzzchick could only find 5 out of
the 20 injected bugs with a 100\% efficacy.
%
%% the hardest one being found only around 13\% of the time after an hour of
%% testing.
%
Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
better than using normal random testing coupled with na\"ive random generators
by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%
However, we believe that a more meticulous evaluation ought to take failure to
find a counterexample as an important metric when comparing PBT tools, not just
mean time to failure --- when one is found.

@@

\citeauthor{lampropoulos2019coverage} demonstrated empirically that \fuzzchick
lies comfortably in the middle ground between using pure random testing with
na\"ive automatically derived random generators and complex manually-written
ones.
%
Their results suggest that CGPT is an appealing technique for finding bugs while
still using a mostly automated workflow.

However, the authors acknowledge that certain parts of its implementation have
room for improvement, especially when it comes to the mutator's design.
%
In tandem with the lack of efficacy observed when replicating the evaluation of
the IFC stack machine case study, these observations led us to consider three
main aspects of \fuzzchick that can be improved upon --- and that constitute the
main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in Section \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in Section \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QSucc, QDisc, gen}}{
    \If{\KwNot Empty(QSucc)}{
      batch $\gets$ Deque(QSucc)\;
      \If{Empty(batch)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        PushFront(QSucc, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \If{\KwNot Empty(QSucc)}{
      batch $\gets$ Deque(QSucc)\;
      \If{Empty(batch)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        PushFront(QSucc, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \lElse{\KwRet Sample(gen)}
  }
\caption{\label{algo:mutagen:pick}\mutagen Seed Selection}
\vspace{1pt}
\end{algorithm}
\end{multicols}
\vspace{-20pt}
\end{figure}

@@

\begin{figure}[b]
\vspace{-15pt}
\newcommand\stacked[2]{\underset{\textstyle #2\mathstrut}{#1}}
\begin{equation}
  \nonumber
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\color{red} \small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\color{blue} \small \texttt{5}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=1\phantom{x}}}{{\color{red} depth=3}}
  \qquad
  \qquad
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{6}}
          child {node {\small \texttt{7}}
          }
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=1cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
        child {node {\color{blue}\small \texttt{6}}
          child {node {\color{blue}\small \texttt{7}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=2\phantom{x}}}{{\color{red} depth=2}}
\end{equation}
\vspace{-10pt}
\caption{\label{fig:mutagen:tracelog}Inserting two new execution traces into the
  internal trace log (represented using brackets).}
\end{figure}

@@

@@

\begin{table}[t]
\footnotesize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed memory alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction when operands are equal \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
\vspace{-25pt}
\end{table}

@@

\begin{table}[b]
\footnotesize
\begin{tabular}{|c|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Category}
& \textbf{Description} \\
\hline
1
& Validator
& Bug
& Invalid memory alignment validation \\
\hline
2
& Validator
& Discrepancy
& Validator accepts blocks returning multiple values  \\
\hline
3
& Interpreter
& Bug
& Instance function invoker silently proceeds after arity mismatch \\
\hline
4
& Interpreter
& Bug
& Allowed out-of-bounds memory access \\
\hline
5
& Interpreter
& Discrepancy
& NaN reinterpretation does not follow reference implementation\\
\hline
\end{tabular}
\caption{\label{table:wasm:bugs}Bugs and discrepancies found by \mutagen in
  \textit{haskell-wasm}.}
\vspace{-20pt}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{tikz/wasm-mttf.tikz}
  \includegraphics[width=\linewidth]{tikz/wasm-fr.tikz}\vspace{-5pt}
  \includegraphics[width=\linewidth]{tikz/wasm-legend.tikz}
  \vspace{-20pt}
  \caption{\label{fig:results:wasm} Comparison between \quickcheck and \mutagen
    across 15 different bugs injected into \textit{haskell-wasm}. }
\vspace{-5pt}
\end{figure}

@@

\begin{table}[H]
\footnotesize
\vspace{-5pt}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multicolumn{1}{|c|}{
  \multirow{2}{*}{Property}}
  & \multicolumn{2}{c|}{\quickcheck}
  & \multicolumn{2}{c|}{\mutagen} \\
  \cline{2-5}
  \multicolumn{1}{|c|}{}
  & \multicolumn{1}{c|}{total}
  & passed
  & \multicolumn{1}{c|}{total}
  & passed \\
\hline
\textit{prop\_validator}
& 41374.92
& 0.00029
& 2243.26
& 519.83\\
\hline
\textit{prop\_interpreter\_i32}
& 134000
& 23.16
& 1815.81
& 351.16 \\
\hline
\end{tabular}
\caption{\label{table:wasm:overhead}Mean number of tests per second executed by
  each tool on the WebAssembly case study.}
\vspace{-20pt}
\end{table}

>>>>

%
%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
%% In this work, we take a close look at \fuzzchick for opportunities to improve.
%
%% Notably, we establish several aspects of \fuzzchick that can be
%% revised.%
%
The approach taken by \fuzzchick leaves considerable room for
improvements in several aspects.%
\footnote{Some of these were already identified by the authors of \fuzzchick in
  their original work.}
%But \fuzzchick suffers several disadvantages.%
%
In particular:
%
\begin{inparaenum}
\item type-preserving mutations, if not done carefully, can become superficial
  and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause large delays if interesting values are
  enqueued frequently and there is no way to prioritize them, and
\item using an heuristic to assign a certain ``mutation energy'' to each
  interesting test case (often referred to as a \emph{power schedule}) requires
  fine tuning and can be hard to generalize to work well under different testing
  scenarios.
\end{inparaenum}
%
%% Moreover, we replicated the Information-Flow Control (IFC) stack machine case
%% study bundled with \fuzzchick and observed a surprising limitation:
%% %
%% after repeating each experiment 30 times, \fuzzchick could only find 5 out of
%% the 20 injected bugs with a 100\% efficacy.
%% %
%% %% the hardest one being found only around 13\% of the time after an hour of
%% %% testing.
%% %
%% Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
%% better than using normal random testing coupled with na\"ive random generators
%% by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%% %
%% However, we believe that a more meticulous evaluation ought to take failure to
%% find a counterexample as an important metric when comparing PBT tools, not just
%% mean time to failure --- when one is found.

@@

\citeauthor{lampropoulos2019coverage} demonstrated empirically that \fuzzchick
lies comfortably in the middle ground between using pure random testing with
na\"ive automatically derived random generators and complex manually-written
ones.
%
Their results suggest that CGPT is an appealing technique for finding bugs while
still using a mostly automated workflow.
%
However, the authors acknowledge that certain parts of its implementation have
room for improvement, especially when it comes to the mutator's design.
%
Moreover, we replicated the Information-Flow Control (IFC) stack machine case
study bundled with \fuzzchick and observed a surprising limitation:
%
\emph{after repeating each experiment 30 times, \fuzzchick could only find 5 out
  of the 20 injected bugs with a 100\% efficacy, the hardest one being found
  only around 13\% of the time after an hour of testing.}
%
Despite this, the authors show that \fuzzchick outperforms normal random testing
coupled with na\"ive random generators by comparing the mean-time-to-failure
(MTTF) against \emph{QuickChick}.
%
In turn, we believe that a more meticulous evaluation ought to take failure to find a
counterexample as an important metric when comparing PBT tools, not just mean
time to failure --- when one is found.
%
%% In tandem with the lack of efficacy observed when replicating the evaluation of
%% the IFC stack machine case study,

All these observations led us to consider three main aspects of \fuzzchick that
can be improved upon --- and that constitute the main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in Section \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in Section \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QSucc, QDisc, gen}}{
    \If{\KwNot Empty(QSucc)}{
      batch $\gets$ Deque(QSucc)\;
      \If{Empty(batch)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        PushFront(QSucc, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \If{\KwNot Empty(QSucc)}{
      batch $\gets$ Deque(QSucc)\;
      \If{Empty(batch)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        PushFront(QSucc, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \lElse{\KwRet Sample(gen)}
  }
\caption{\label{algo:mutagen:pick}\mutagen Seed Selection}
\vspace{1pt}
\end{algorithm}
\end{multicols}
\vspace{-10pt}
\end{figure}

@@

\begin{figure}[t]
\vspace{-15pt}
\newcommand\stacked[2]{\underset{\textstyle #2\mathstrut}{#1}}
\begin{equation}
  \nonumber
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\color{red} \small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\color{blue} \small \texttt{5}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=1\phantom{x}}}{{\color{red} depth=3}}
  \qquad
  \qquad
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{6}}
          child {node {\small \texttt{7}}
          }
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=1cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
        child {node {\color{blue}\small \texttt{6}}
          child {node {\color{blue}\small \texttt{7}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=2\phantom{x}}}{{\color{red} depth=2}}
\end{equation}
\vspace{-10pt}
\caption{\label{fig:mutagen:tracelog}Inserting two new execution traces into the
  internal trace log (represented using brackets).}
\end{figure}

@@

\begin{figure}[b]
\vspace{-15pt}
\begin{multicols}{2}

@@

\begin{table}[t]
\footnotesize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed memory alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction when operands are equal \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
\vspace{-15pt}
\end{table}

@@

\begin{table}[t]
\footnotesize
\begin{tabular}{|c|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Category}
& \textbf{Description} \\
\hline
1
& Validator
& Bug
& Invalid memory alignment validation \\
\hline
2
& Validator
& Discrepancy
& Validator accepts blocks returning multiple values  \\
\hline
3
& Interpreter
& Bug
& Instance function invoker silently proceeds after arity mismatch \\
\hline
4
& Interpreter
& Bug
& Allowed out-of-bounds memory access \\
\hline
5
& Interpreter
& Discrepancy
& NaN reinterpretation does not follow reference implementation\\
\hline
\end{tabular}
\caption{\label{table:wasm:bugs}Bugs and discrepancies found by \mutagen in
  \textit{haskell-wasm}.}
\vspace{-20pt}
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{tikz/wasm-mttf.tikz}
  \includegraphics[width=\linewidth]{tikz/wasm-fr.tikz}\vspace{-5pt}
  \includegraphics[width=\linewidth]{tikz/wasm-legend.tikz}
  \vspace{-20pt}
  \caption{\label{fig:results:wasm} Comparison between \quickcheck and \mutagen
    across 15 different bugs injected into \textit{haskell-wasm}. }
%% \vspace{-5pt}
\end{figure}

@@

\begin{table}[H]
\footnotesize
%% \vspace{-5pt}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multicolumn{1}{|c|}{
  \multirow{2}{*}{Property}}
  & \multicolumn{2}{c|}{\quickcheck}
  & \multicolumn{2}{c|}{\mutagen} \\
  \cline{2-5}
  \multicolumn{1}{|c|}{}
  & \multicolumn{1}{c|}{total}
  & passed
  & \multicolumn{1}{c|}{total}
  & passed \\
\hline
\textit{prop\_validator}
& 41374.92
& 0.00029
& 2243.26
& 519.83\\
\hline
\textit{prop\_interpreter\_i32}
& 134000
& 23.16
& 1815.81
& 351.16 \\
\hline
\end{tabular}
\caption{\label{table:wasm:overhead}Mean number of tests per second executed by
  each tool on the WebAssembly case study.}
\vspace{-10pt}
\end{table}
